### 2015-06-04 doc/thesis/evaluation.tex

Table \ref{tab:evalbycase} summarizes our evaluation results for each
of the case studies that we used Typed Lua for typing their members.
An exported member is any Lua value that a module might export.
We split the members of each case study into four categories:
\emph{easy}, \emph{poly}, \emph{over}, and \emph{hard}.
The \emph{easy} category shows the percentage of members that
we could give a precise static type.
The \emph{poly} category shows the percentage of members that
we made minimal use of the dynamic type as a replacement for the
lack of type parameters.
The \emph{over} category shows the percentage of members that
we gave a static type that is not precise enough,
as these members are overloaded functions that require intersection
types to describe their precise static types.
The \emph{hard} category shows the percentage of members that
we could not give a precise static type.
The last column of the table shows the total number of members
of each case study.
The results that we obtained for the \emph{easy} category give
a lower bound on how much static type safety
we could add to each one of our case studies.

@@

Even though the \texttt{math} module looks straightforward to type,
the \emph{over} category shows that it includes several overloaded functions.
For instance, the functions \texttt{math.abs} and \texttt{math.random}
are in this category.

In the case of \texttt{math.abs},
it has two different types: \texttt{(integer) -> (integer)} and
\texttt{(number) -> (number)}.
Even though we gave this function the more general type \texttt{(number) -> (number)},
it is not precise enough because the return type is always \texttt{number}
independently of the argument type.
In other words, the return type should be \texttt{integer} when
the argument type is also \texttt{integer}, and the return type
should be \texttt{number} when the argument type is also \texttt{number}.

In the case of \texttt{math.random}, it also has two different types:
\texttt{() -> (number)} and \texttt{(integer, integer?) -> (integer)}.
This means that the type of \texttt{math.random} depends on the calling arity.
Calling \texttt{math.random()} returns a random floating point
between \texttt{0} and \texttt{1}.
Calling \texttt{math.random(10)} is equivalent to \texttt{math.random(1,10)},
and returns an integer between this interval.
Like \texttt{table.insert}, this function also does not follow the
semantics of Lua on discarding extra arguments,
and generates a run-time error whenever we pass more than two arguments.

@@

Even tough it was straightforward to type the MD5 library,
we found a little difference between its documentation and its behavior.
The documentation suggests that the type of the function \texttt{md5.update}
is \texttt{(md5\string_context, string) -> (md5\string_context)},
though there is a call to this function in the test script that passes
an extra string argument.
Reading the source code, we found that its actual type is
\texttt{(md5\string_context, string*) -> (md5\string_context)},
that is, we can pass zero or more strings to \texttt{md5.update}.

@@

We could give precise static types to most of the members
in the \texttt{socket} module, which is the C core.
However, this module includes some functions that
we could not give a precise static type because they
rely on reflection, as the \emph{hard} category shows.
For instance, \texttt{socket.skip} is a function that is
in this category.
We can use this function to choose the number of
values that we want to return.
As an example, calling \texttt{socket.skip(1, nil, "hello")}
returns only the string \texttt{"hello"}, because \texttt{1} indicates
that we do not want to return the first value.
Passing a negative number to \texttt{socket.skip} can be
dangerous, as it returns anything that might be in the stack.
As an example, calling \texttt{socket.skip(-1, nil, "hello")}
returns the tuple \texttt{(-1, nil, "hello")}, because \texttt{-1} makes
\texttt{socket.skip} does not skip any values.
As another example, \texttt{f = socket.skip(-2)} assigns \texttt{socket.skip}
to \texttt{f}, as \texttt{-2} gets \texttt{socket.skip} from the stack.
Our type system cannot handle the type of negative numbers, as
this requires more complex types such as the refinement types from
hybrid type checking \citep{flanagan2006htc}.

@@

The function \texttt{http.request} downloads data from a given URL,
which can be either a string or a table.
More precisely, \texttt{http.request(url, body)} returns the tuple
\texttt{(string, number, \{string:string\}, number) | (nil, string)}
if \texttt{url} is a string and \texttt{body} is another string or \texttt{nil},
but it returns the tuple
\texttt{(number, number, \{string:string\}, number) | (nil, string)}
if \texttt{url} is a table and \texttt{body} is \texttt{nil}.

@@

The \texttt{tllexer} module defines common lexical rules for
the Typed Lua parser and the description file parser.
This module is hard to type because it uses LPeg \citep{lpeg,ierusalimschy2009lpeg}
patterns, and LPeg uses overloaded arithmetic operators to
build LPeg patterns. 
Even though LPeg is the third most popular Lua module,
we cannot not precisely type LPeg patterns because our
type system still does not support overloading arithmetic
operators.
In the \texttt{tllexer} module, we could only give precise static types
to two error reporting functions that it exports.

>>>>

Table \ref{tab:evalbycase} summarizes our evaluation results for each
of the case studies that we used Typed Lua for typing their members.
An exported member is any Lua value that a module might export.
We split the members of each case study into four categories:
\emph{easy}, \emph{poly}, \emph{over}, and \emph{hard}.
In the next four paragraphs we explain each category in more detail.
The last column of the table shows the total number of members
of each case study.

The \emph{easy} category shows the percentage of members that
we could give a precise static type.
For instance, the function \texttt{string.len} from the Lua
standard library is in this category because we could use
Typed Lua to describe its type: \texttt{(string) -> (integer)}.
This function returns the length of a given string.
Note that the results that we obtained for this category give
a lower bound on how much static type safety
we could add to each one of our case studies.

The \emph{poly} category shows the percentage of members that
we made minimal use of the dynamic type as a replacement for the
lack of type parameters.
For instance, the function \texttt{table.sort} from the Lua
standard library is in this category because it is a generic function.
This function sorts a given list of elements, which is a generic list.
However, we had to assign to this function the type
\texttt{(\{any\}, nil|(any, any) -> (boolean)) -> ()} because
Typed Lua does not have parametric polymorphism.

The \emph{over} category shows the percentage of members that
we gave a static type that is not precise enough,
as these members are overloaded functions that require intersection
types to describe their precise static types.
For instance, the function \texttt{math.abs} from the Lua
standard library is in this category because it has two types:
\texttt{(integer) -> (integer)} and \texttt{(number) -> (number)}.
This function returns the absolute value of a given number,
which can be either integer or float.
Even though we gave this function the more general type \texttt{(number) -> (number)},
it is not precise enough because the return type is always \texttt{number}
independently of the argument type.
In other words, the return type should be \texttt{integer} when
the argument type is also \texttt{integer}, and the return type
should be \texttt{number} when the argument type is also \texttt{number}.
However, we cannot give such a precise type to this function because
Typed Lua does not support overloaded functions.

The \emph{hard} category shows the percentage of members that
we could not give a precise static type.
For instance, the function \texttt{assert} from the Lua
standard library is in this category because it can return any number
of values of any type, being difficult to type this behavior.
Still, we gave this function the type \texttt{(value*) -> (any*)}.

@@

Even though the \texttt{math} module looks straightforward to type,
the \emph{over} category shows that it includes several overloaded functions.
For instance, the function \texttt{math.random} is in this category
because it has two different types:
\texttt{() -> (number)} and \texttt{(integer, integer?) -> (integer)}.
This means that the type of \texttt{math.random} depends on the calling arity.
Calling \texttt{math.random()} returns a random floating point
between \texttt{0} and \texttt{1}.
Calling \texttt{math.random(10)} is equivalent to \texttt{math.random(1,10)},
and returns an integer between this interval.
Like \texttt{table.insert}, this function also does not follow the
semantics of Lua on discarding extra arguments,
and generates a run-time error whenever we pass more than two arguments.

@@

Even tough it was straightforward to type the MD5 library,
we found a little difference between its documentation and its behavior.
The documentation suggests that the type of \texttt{md5.update}
is \texttt{(md5\string_context, string) -> (md5\string_context)},
though there is a call to this function in the test script that passes
an extra string argument.
Reading the source code, we found that its actual type is
\texttt{(md5\string_context, string*) -> (md5\string_context)},
that is, we can pass zero or more strings to \texttt{md5.update}.

@@

We could give precise static types to most of the members
in the \texttt{socket} module, which is the C core.
However, this module includes some functions that
we could not give a precise static type because they
rely on reflection, as the \emph{hard} category shows.
For instance, \texttt{socket.skip} is a function that is
in this category.
We can use this function to choose the number of
values that we want to return.
As an example, calling \texttt{socket.skip(1, nil, "hello")}
returns only the string \texttt{"hello"}, because \texttt{1} indicates
that we do not want to return the first value.
Passing a negative number to \texttt{socket.skip} can be
dangerous, as it returns anything that might be in the stack.
As an example, calling \texttt{socket.skip(-1, nil, "hello")}
returns the tuple \texttt{(-1, nil, "hello")}, because \texttt{-1} makes
\texttt{socket.skip} not skip any values.
As another example, the code \texttt{f = socket.skip(-2)} assigns \texttt{socket.skip}
to \texttt{f}, as \texttt{-2} gets \texttt{socket.skip} from the stack.
Our type system cannot handle the type of negative numbers, as
this requires more complex types such as the refinement types from
hybrid type checking \cite{flanagan2006htc}.

@@

The function \texttt{http.request} downloads data from a given URL,
which can be either a string or a table.
More precisely, \texttt{http.request(url, body)} returns the tuple type
\texttt{(string, number, \{string:string\}, number) | (nil, string)}
if \texttt{url} is a string and \texttt{body} is another string or \texttt{nil},
but it returns the tuple type
\texttt{(number, number, \{string:string\}, number) | (nil, string)}
if \texttt{url} is a table and \texttt{body} is \texttt{nil}.

@@

The \texttt{tllexer} module defines common lexical rules for
the Typed Lua parser and the description file parser.
This module is hard to type because it uses LPeg \cite{lpeg,ierusalimschy2009lpeg}
patterns, and LPeg uses overloaded arithmetic operators to
build LPeg patterns. 
Even though LPeg is the third most popular Lua module,
we cannot precisely type LPeg patterns because our
type system still does not support overloading arithmetic
operators.
In the \texttt{tllexer} module, we could only give precise static types
to two error reporting functions that it exports.

### 2015-06-04 doc/thesis/glossary.tex

\item[closed table type] It is the type of table annotations in Typed Lua.
A closed table type does not provide any guarantees that about keys
with types not listed in the table type, and it also does not
allow adding fields to the table type.

@@

\item[consistency] Gradual typing uses the consistency relation instead of type equality
to check the interaction among the dynamic type and other types.
This relation allows us combining dynamic and static typing in the
same language, but still catching static type errors.
The consistency relation is reflexive and symmetric, but it is not transitive.

\item[consistent-subtyping] It is a relation that combines consistency and subtyping, allowing
the definition of gradual type systems for object-oriented languages.
The consistent-subtyping relation is reflexive and symmetric, but it is not transitive.

@@

\item[depth subtyping] It allows the supertype to change the type of individual fields in the subtype.

\item[dynamic type] It is a type that allows combining dynamic and static typing in the same code.
It is neither the bottom nor the top type in the subtyping relation, but
a subtype only of itself.
Gradual typing uses the dynamic type along with the consistency relation to
identify where run-time casts should be inserted to prevent that dynamically
typed code violates statically typed code.

\item[filter type] It is a type that allows Typed Lua to discriminate and restore the type of
local variables inside control flow statements.

\item[fixed table type] It is a table type that does not allow width subtyping to make single
inheritance safe in Typed Lua, and it also does not allow adding fields to
the table type.

@@

\item[gradual type system] It is a type system that uses the consistency relation instead of type equality
to check the interaction among the dynamic type and other types.

\item[gradual typing] It is an approach that uses a gradual type system to allow static and dynamic
typing in the same code, but inserting run-time checks between statistically
typed and dynamically typed code.
This run-time checks ensure that dynamically typed code does not violate
statically typed code.

@@

\item[open table type] It is the type of the tables with keys that do not inhabit one of
the table's key types, have at least one alias, and allows adding
fields to the table type.

\item[optional type system] It is a type system that allows combining static and dynamic typing in the same
language, but without affecting the run-time semantics.

\item[projection environment] It is an environment that Typed Lua uses to assign projection variables to
second-level types.

\item[projection type] It is a type that allows Typed Lua to discriminate the type of local variables
that have a dependency relation, as they include a projection variable and an
index that allow Typed Lua to project unions of second-level types into
unions of first-level types.

@@

\item[self-like delegation] It is a technique to implement inheritance in dynamically typed languages
through prototype objects.

@@

\item[table refinement] It is an operation that allows changing a table type to include new fields or
to specialize existing fields.

@@

\item[type tag] In dynamically typed languages a type tag describes the type of a value during
run-time.

\item[unique table type] It is the type of the tables with keys that do not inhabit one of
the table's key types, does not have any alias, and allow adding
fields to the table type.
It is also the table type that describes the type of the table constructor.

>>>>

\item[closed table type] It is the type of table annotations in Typed Lua.
A closed table type does not provide any guarantees about keys
with types not listed in the table type.
It also does not allow table refinement to add fields to the table type.

@@

\item[consistency] Gradual typing uses the consistency relation to check the interaction
among the dynamic type and other types.
This relation allows us combining dynamic and static typing in the
same language, but still catching static type errors.
The consistency relation is reflexive and symmetric, but it is not transitive
to prevent that casts from a static type to the dynamic type can be
combined with casts from the dynamic type to another static type.

\item[consistent-subtyping] It is a relation that combines consistency and subtyping, allowing
the definition of gradual type systems for object-oriented languages.
Like the consistency relation, it is reflexive and symmetric, but it is not transitive.

@@

\item[depth subtyping] It allows the supertype to vary the type of individual fields in the subtype.

\item[dynamic type] It is a type that allows combining dynamic and static typing in the same code.
It is neither the bottom nor the top type in the subtyping relation, but
a subtype only of itself.
Gradual typing uses the dynamic type along with the consistency relation to
identify the parts of the code where run-time casts should be inserted to
prevent that dynamically typed code violates statically typed code.

\item[filter type] It is a type that allows Typed Lua to discriminate the type of local variables
inside control flow statements, as these variables are bound to unions of
first-level types.

\item[fixed table type] It is the type of classes in Typed Lua.
It is a table type that does not allow width subtyping to make single
inheritance safe.
It also does not allow table refinement to add fields to the table type.

@@

\item[gradual type system] It is a type system that uses the consistency relation instead of type equality
to perform static type checking.

\item[gradual typing] It is an approach that uses a gradual type system to allow static and dynamic
typing in the same code, but inserting run-time checks between statically
typed and dynamically typed code.
These run-time checks ensure that dynamically typed code does not violate
statically typed code.
More precisely, gradual typing allows programmers to change between
dynamic and static typing without changing the dynamic or the static behavior
of the program.

@@

\item[open table type] It is the type of the tables with keys that do not inhabit one of
the table's key types, and have at least one alias.
It also allows table refinement to add fields to the table type.

\item[optional type system] It is a type system that allows combining static and dynamic typing in the same
language, but without affecting the run-time semantics.
This means that the programmer has the option to use or not use its
static analysis to check for static type errors, as these errors
will be caught by the run-time semantics anyway.

\item[projection environment] It is an environment that Typed Lua uses to handle unions of second-level
types that are bound to projection types.

\item[projection type] It is a type that allows Typed Lua to discriminate the type of local variables
that have a dependency relation, as they project an union of second-level types
into unions of first-level types.
These unions of second-level types can describe the type of several local variables.
A projection type uses an index to define which components of a union of second-level
types should be used to project the union of first-level types of a local variable.
When programmers discriminate a projection type, they are discriminating the
union of second-level types that is bound to this projection type, affecting
all variables that bind to the same projection type, which also bind to the
same union of second-level types.

@@

\item[self-like delegation] It is a technique to implement inheritance in dynamically typed languages
through prototype objects. 
In this technique, when an object tries to access an attribute that is not
present, it looks for this attribute in its parent object.

@@

\item[table refinement] It is an operation from Typed Lua that allows programmers to change a table type
to include new fields or to specialize existing fields.

@@

\item[type tag] In dynamically typed languages, a type tag describes the type of a value during run-time.

\item[unique table type] It is the table type that describes the type of a table constructor.
It is the type of the tables with keys that do not inhabit one of
the table's key types, and that does not have any alias.
It also allows table refinement to add fields to the table type.

### 2015-06-04 doc/thesis/related.tex
Fixing typos

In this chapter we review related work as a way to identify the novel
features of our type system, what are the features that it shares with
other type systems, and whether we can incorporate some interesting
features from other type systems.
We split this chapter into two sections: in the first section we review
other Lua projects, while in the second section we review other
projects that are not related to Lua.

@@

TypeScript \citep{typescript} is a JavaScript extension
that includes optional type annotations and class-based
object-oriented programming.
It also uses local type inference to deduce the type
of unannotated expressions.
The type system of TypeScript is structural, based
on subtyping, and supports generics.
It includes the dynamic type, primitive types, union types,
function types, array types, tuple types, recursive types, and
object types.
Unlike Typed Lua, TypeScript uses arrays to represent variadic
functions and multiple return values.

Even though TypeScript is unsound by design,
\citet{bierman2014typescript} shows how to make TypeScript sound.
They use a reduced core of TypeScript to formalize a
sound type system for TypeScript, but also to formalize
its current unsound type system.

TeJaS \citep{lerner2013tejas} is a framework for the construction of
different type systems for JavaScript.
The authors created a base type system for JavaScript with
extensible typing rules that allow the experimentation of
different static analysis.
They used TeJaS to create a type system that simulates the
type system of TypeScript.

\citet{politz2012semantics} proposes semantics and types for objects
with first-class member names, a well-known feature from scripting languages.
Their type system uses string patterns to describe the members of
an object, and define a complex subtyping relation to validate
these patterns.
They also provide an implementation of their system to JavaScript.

Gradualtalk \citep{allende2013gts} is a Smalltalk dialect that
supports gradual typing.
The type system combines nominal and structural typing.
It includes function types, union types, structural types,
nominal types, a self type, and parametric polymorphism.
The type system also relies on subtyping and consistent-subtyping.

Gradualtalk inserts run-time checks that ensure dynamically
typed code does not violate statically typed code.
\citet{allende2013cis} perform a carefull evaluation about
cast insertion in Gradualtalk.
They report that usually cast insertions impact on execution
performance, so Gradualtalk also has an option that allows
programmers to turn them off, downgrading Gradualtalk
to an optional type system.

Reticulated Python \citep{vitousek2014deg} is a Python compiler
that supports gradual typing.
The type system is structural and based on subtyping.
It includes base types, the dynamic type, list types,
dictionary types, tuple types, function types, set types,
object types, class types, and recursive types.
It includes class and object types to differentiate the
type of class declarations and instances, respectively.
It also uses local type inference.
Besides static type checking, Reticulated Python also introduces
three different approaches for inserting run-time assertions.

Mypy \citep{mypy} is an optional type system for Python.
The type system of mypy is similar to the type system of
Reticulated Python, but mypy does not insert run-time checks
and it has parametric polymorphism.
In contrast, Reticulated Python can type variadic functions,
but mypy cannot.
Recently, Guido van Rossum, Python's author, proposed a
standard syntax for type annotations in Python \citep{PEP483}
that is extremely inspired by mypy \citep{PEP484}.
The main goal of this proposal is to make easier building
static analysis tools for Python.
Typing \citep{typing} is a tool that is being developed to
implement this proposal.

Hack \citep{hack} is a new programming language that runs on the
Hip Hop Virtual Machine (HHVM).
The HHVM is a virtual machine that executes Hack and PHP programs.
We can view Hack as an extension to PHP that combines static and
dynamic typing.
The type system of Hack includes generics, nullable types, collections,
and function types.

The Ruby Type Checker \citep{ren2013rtc} is a library that
performs type checking during run-time.
The library provides type annotations that the programmer
can use on classes and methods.
Its type system includes nominal types, union types,
intersection types, method types, parametric polymorphism,
and type casts.

Grace \citep{black2013sg} is an object-oriented language
with optional typing.
Grace is not a dynamically typed language that has been
extended with an optional type system, but a language
that has been designed from scratch to have both
static and dynamic typing.
\citet{homer2013modules} explores some useful patterns
that derive from Grace's use of objects as modules
and its brand of optional structural typing, which
can also be expressed with Typed Lua's modules as tables.

>>>>

In this chapter we review related work, and we split it into two sections:
in the first section we review other Lua projects,
while in the second section we review other projects that are not related to Lua.

@@

TypeScript \cite{typescript} is a JavaScript extension
that includes optional type annotations and class-based
object-oriented programming.
It also uses local type inference to deduce the type
of unannotated expressions.
The type system of TypeScript is structural, based
on subtyping, and supports generics.
It includes the dynamic type, primitive types, union types,
function types, array types, tuple types, recursive types, and
object types.
Unlike Typed Lua, TypeScript uses arrays to represent variadic
functions and multiple return values.

Even though TypeScript is unsound by design,
Bierman et al. \cite{bierman2014typescript} shows how to
make TypeScript sound.
They use a reduced core of TypeScript to formalize a
sound type system for TypeScript, but also to formalize
its current unsound type system.

TeJaS \cite{lerner2013tejas} is a framework for the construction of
different type systems for JavaScript.
The authors created a base type system for JavaScript with
extensible typing rules that allow the experimentation of
different static analysis.
They used TeJaS to create a type system that simulates the
type system of TypeScript.

Politz et al. \cite{politz2012semantics} proposes semantics
and types for objects with first-class member names, a well-known
feature from scripting languages.
Their type system uses string patterns to describe the members of
an object, and define a complex subtyping relation to validate
these patterns.
They also provide an implementation of their system to JavaScript.

Gradualtalk \cite{allende2013gts} is a Smalltalk dialect that
supports gradual typing.
The type system combines nominal and structural typing.
It includes function types, union types, structural types,
nominal types, a self type, and parametric polymorphism.
The type system also relies on subtyping and consistent-subtyping.

Gradualtalk inserts run-time checks that ensure dynamically
typed code does not violate statically typed code.
Allende et al. \cite{allende2013cis} perform a careful
evaluation about cast insertion in Gradualtalk.
They report that usually cast insertions impact on execution
performance, so Gradualtalk also has an option that allows
programmers to turn them off, downgrading Gradualtalk
to an optional type system.

Reticulated Python \cite{vitousek2014deg} is a Python compiler
that supports gradual typing.
The type system is structural and based on subtyping.
It includes base types, the dynamic type, list types,
dictionary types, tuple types, function types, set types,
object types, class types, and recursive types.
It includes class and object types to differentiate the
type of class declarations and instances, respectively.
It also uses local type inference.
Besides static type checking, Reticulated Python also introduces
three different approaches for inserting run-time assertions.

Mypy \cite{mypy} is an optional type system for Python.
The type system of mypy is similar to the type system of
Reticulated Python, but mypy does not insert run-time checks
and it has parametric polymorphism.
In contrast, Reticulated Python can type variadic functions,
but mypy cannot.
Recently, Guido van Rossum, Python's author, proposed a
standard syntax for type annotations in Python \cite{PEP483}
that is extremely inspired by mypy \cite{PEP484}.
The main goal of this proposal is to make easier building
static analysis tools for Python.
Typing \cite{typing} is a tool that is being developed to
implement this proposal.

Hack \cite{hack} is a new programming language that runs on the
Hip Hop Virtual Machine (HHVM).
The HHVM is a virtual machine that executes Hack and PHP programs.
We can view Hack as an extension to PHP that combines static and
dynamic typing.
The type system of Hack includes generics, nullable types, collections,
and function types.

The Ruby Type Checker \cite{ren2013rtc} is a library that
performs type checking during run-time.
The library provides type annotations that the programmer
can use on classes and methods.
Its type system includes nominal types, union types,
intersection types, method types, parametric polymorphism,
and type casts.

Grace \cite{black2013sg} is an object-oriented language
with optional typing.
Grace is not a dynamically typed language that has been
extended with an optional type system, but a language
that has been designed from scratch to have both
static and dynamic typing.
Homer et al. \cite{homer2013modules} explores some
useful patterns that derive from Grace's use of objects as modules
and its brand of optional structural typing, which
can also be expressed with Typed Lua's modules as tables.

### 2015-06-04 doc/thesis/system.tex

Back to our example, the table constructor
$\{ [1] = ``x", [2] = ``y", [3] = ``z" \}$ has type
$\{1:``x", 2:``y", 3:``z"\}$ through rule \textsc{T-CONSTRUCTOR2}.

As another example, the table constructor $\{[``x"] = 1, [``y"] = \{[``z"] = 2\}\}$
has type $\{``x":1, ``y":\{``z":2\}_{closed}\}_{unique}$ through rule
\textsc{T-CONSTRUCTOR2}.
The inner table is \emph{closed} to prevent the creation of unsafe aliases.

After presenting some typing rules of the table constructor,
we start the discussion of the rules that define the most
unusual feature of our type system: the refinement of table types.
The first kind of refinement allows programmers to add new
fields to \emph{unique} or \emph{open} table types through
field assignment.
For instance, in Section \ref{sec:tables} we presented the
following example:
\begin{verbatim}

    local person = {}
    person.firstname = "Lou"
    person.lastname = "Reed"
\end{verbatim}

@@

This rule uses the auxiliary function \emph{infer} to get the
most general first-level types of each variable that should be introduced in
the type environment for type checking $s$.
After type checking the statement $s$, rule \textsc{T-LOCAL2} produces a
new type environment $\env_{3}$ without the variables that it introduced
before type checking $s$.
We can define \emph{infer} as follows:
\begin{align*}
infer(T_{1} \times ... \times T_{n}{*}, i) & = \left\{
\begin{array}{ll}
general(T_{i}) & \text{if $i < n$}\\
general(T_{n} \cup \Nil) & \text{if $i >= n$}
\end{array} \right.
\\ \\
general(\False) & = \Boolean\\
general(\True) & = \Boolean\\
general({\it int}) & = \Integer\\
general({\it float}) & = \Number\\
general({\it string}) & = \String\\
general(T_{1} \cup T_{2}) & = general(T_{1}) \cup general(T_{2})\\
general(S_{1} \rightarrow S_{2}) & = general2(S_{1}) \rightarrow general2(S_{2})\\
general(\{K_{1}{:}V_{1}, ..., K_{n}{:}V_{n}\}_{tag}) & = \{K_{1}{:}general(V_{1}), ..., K_{n}{:}general(V_{n})\}_{tag}\\
general(\mu x.T) & = \mu x.general(T)\\
general(T) & = T\\
\\
general2(\Void) & = \Void\\
general2(T{*}) & = general(T){*}\\
general2(T \times P) & = general(T) \times general2(P)\\
general2(S_{1} \sqcup S_{2}) & = general2(S_{1}) \sqcup general2(S_{2})
\end{align*}

>>>>

Back to our example, the constructor
$\{ [1] = ``x", [2] = ``y", [3] = ``z" \}$ has type
$\{1:``x", 2:``y", 3:``z"\}_{unique}$ through rule \textsc{T-CONSTRUCTOR2}.

As another example, the constructor $\{[``x"] = 1, [``y"] = \{[``z"] = 2\}\}$
has type $\{``x":1, ``y":\{``z":2\}_{closed}\}_{unique}$ through rule
\textsc{T-CONSTRUCTOR2}.
The inner table is \emph{closed} to prevent the creation of unsafe aliases.

After presenting some typing rules of the table constructor,
we start the discussion of the rules that define the most
unusual feature of our type system: the refinement of table types.
The first kind of refinement allows programmers to add new
fields to \emph{unique} or \emph{open} table types through
field assignment.
For instance, in Section \ref{sec:tables} we presented the
following example:
\begin{verbatim}
    local person = {}
    person.firstname = "Lou"
    person.lastname = "Reed"
\end{verbatim}

@@

This rule uses the auxiliary function \emph{infer} to get the
most general first-level types of each variable that should be introduced in
the type environment for type checking $s$.
After type checking the statement $s$, rule \textsc{T-LOCAL2} produces a
new type environment $\env_{3}$ without the variables that it introduced
before type checking $s$.
We can define \emph{infer} as follows:
\begin{align*}
infer(T_{1} \times ... \times T_{n}{*}, i) & = \left\{
\begin{array}{ll}
general(T_{i}) & \text{if $i < n$}\\
general(T_{n} \cup \Nil) & \text{if $i >= n$}
\end{array} \right.
\\ 
general(\False) & = \Boolean\\
general(\True) & = \Boolean\\
general({\it int}) & = \Integer\\
general({\it float}) & = \Number\\
general({\it string}) & = \String\\
general(T_{1} \cup T_{2}) & = general(T_{1}) \cup general(T_{2})\\
general(S_{1} \rightarrow S_{2}) & = general2(S_{1}) \rightarrow general2(S_{2})\\
general(\{K_{1}{:}V_{1}, ..., K_{n}{:}V_{n}\}_{tag}) & = \{K_{1}{:}general(V_{1}), ..., K_{n}{:}general(V_{n})\}_{tag}\\
general(\mu x.T) & = \mu x.general(T)\\
general(T) & = T\\
\end{align*}

\begin{align*}
general2(\Void) & = \Void\\
general2(T{*}) & = general(T){*}\\
general2(T \times P) & = general(T) \times general2(P)\\
general2(S_{1} \sqcup S_{2}) & = general2(S_{1}) \sqcup general2(S_{2})
\end{align*}
